---
title: "RAbHIT: R Antibody Haplotype Infrence Tool"
author: "Moriah Gidoni, Ayelet Peres, Gur Yaari"
date: "Last modified `r Sys.Date()`"
output:
  pdf_document:
    dev: pdf
    fig_height: 20
    fig_width: 15
    highlight: pygments
    toc: yes
  html_document:
    fig_height: 20
    fig_width: 15
    highlight: pygments
    theme: readable
    toc: yes
  md_document:
    fig_height: 20
    fig_width: 15
    preserve_yaml: no
    toc: yes
geometry: margin=1in
fontsize: 11pt
vignette: >
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteIndexEntry{Haplotype infrence} 
  %\usepackage[utf8]{inputenc}
---

## Introduction 

Analysis of antibody repertoires by high throughput sequencing is of major importance in understanding adaptive immune responses. 
Our knowledge of variations in the genomic loci encoding antibody genes is incomplete, mostly due to technical difficulties in aligning short reads to these highly repetitive loci. 
The partial knowledge results in conflicting \textit{V-D-J} gene assignments between different algorithms, and biased genotype and haplotype inference. Previous studies have shown that haplotypes can be inferred by taking advantage of \textit{IGHJ6} heterozygosity, observed in approximately one third of the population.

Here we provide a robust novel method for determining \textit{V-D-J} haplotypes by adapting a Bayesian framework, **Haplotyper**. 
Our method extends haplotype inference to \textit{IGHD}- and \textit{IGHV}-based analysis, thereby enabling inference of complex genetic events like deletions and copy number variations in the entire population. It calculates a Bayes factor, a number that indicates the certainty level of the inference, for each haplotyped gene. 

More details can be found here:

[Mosaic deletion patterns of the human antibody heavy chain gene locus as revealed by Bayesian haplotyping](https://doi.org/10.1101/314476).


## Input

HaplotypeR requires two main inputs:

1. Pre-processed antibody repertoire sequencing data with heterozygosity in at least one gene
2. Database of germline gene sequences

Antibody repertoire sequencing data is in a data frame format. Each row represents a unique 
observation and columns represent data about that observation. The names of the required columns are provided below along with a short description.

Column Name           | Description 
----------------------|---------------------------------------------------------
`SUBJECT`             | Subject name
`V_CALL`              | (Comma separated) name(s) of the nearest V allele(s) (IMGT format)
`D_CALL`              | (Comma separated) name(s) of the nearest J allele(s) 
`J_CALL`              | (Comma separated) name(s) of the nearest J allele(s) 


An example dataset is provided with the `rabhit` package. It contains unique naive b-cell
sequences, from a single individual.

The database of germline sequences should be provided in FASTA format with 
sequences gapped according to the IMGT numbering scheme ([[3]][3]). IGHV alleles in
the IMGT database (build 201408-4) are provided with this package. (object name)

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(rabhit)
# Load example sequence data and example germline database
data(sample_db, HVGERM, HDGERM)
```
### Pre-processing of the data

One can use TIgGER functions to filter the data.....It is important to note which cell types/ exp. protocol....one representative from each clone (see SHAZAM...).

## Running RAbHIT

The functions provided by this package can be used to perform any combination of
the following:

1. Infer haplotype by anchor gene
2. Infer D/J single chromosome deletions by the V pooled approach
3. Infer double chromosome deletion by relative gene usage
4. Graphical output of the inferred haplotype 
5. Graphical output of the inferred deletion

### Infer haplotype by anchor gene

An individual's haplotype can be inferred using the functions `createFullHaplotype`.
The function infers the haplotype based on the provided anchor gene. Using this function a contingency table is created for each gene, *from which strand is inferred for each allele*.
The user can set the anchor gene for haplotyping as well as the column for which a haplotype should be inferred.

Prior to haplotyping, it is recommended to run TIgGER on the data, to detect new alleles and construct a genotype [link to docs].

```{r, eval=TRUE, warning=FALSE}
# Infered haplotype summary table
haplo_db <- createFullHaplotype(sample_db,toHap_col=c("V_CALL","D_CALL"),
                            hapBy_col="J_CALL",hapBy="IGHJ6",toHap_GERM=c(HVGERM,HDGERM))

head(haplo_db)
```

```{r, eval=TRUE, warning=FALSE}
# Plot the haplotype
plotHaplotype(haplo_db)
```

```{r, eval=TRUE, warning=FALSE}
# Plot interactive haplotype plot
p <- plotHaplotype(haplo_db,html_output = T)
#save plot to html output
htmlwidgets::saveWidget(p, "haplotype.html",selfcontained = T)
```

### Haplotype inference deletion heatmap

For a group of individuals, the deletions detected by the haplotyping process can be visualized with `deletionHeatmap`.
The function creatre a heatmap of the deletion infered and colors them by the certainty level (lK).

```{r, eval=TRUE, warning=FALSE,fig.height=12,fig.width=15}
# Load example sequence data
data(samples_db)
# Infered haplotype summary table for multiple subjects
haplo_db <- createFullHaplotype(samples_db,toHap_col=c("V_CALL","D_CALL"),
                            hapBy_col="J_CALL",hapBy="IGHJ6",toHap_GERM=c(HVGERM,HDGERM))
# plot deletion heatmap
deletionHeatmap(haplo_db)
```


### Infering D/J single chromosome deletion by V pooled approach

Since V gene heterozygosity is extremely common,
using V genes as anchors for haplotype inference could dramatically 
increase the number of people for which D haplotype can be inferred.
However, reliable haplotype inference using V genes as 
anchors requires a much greater sequencing depth than haplotype inference using J6 gene as an anchor.

The haplotper package offeres a solution to  overcome the low number of sequences that connect a given V-D allele pair.
The package function applies an aggregation approach, in which information from several V heterozygous genes can be combined to infer D gene
deletions.

The `deletionsByVpooled` function uses the V pooled approach to detect single chromosomal deletion for D and J.


```{r, eval=TRUE, warning=FALSE}
# Infered deletion summary table
del_db <- deletionsByVpooled(samples_db)
head(del_db)
```

For visualzing the deletion detected by `deletionsByVpooled`, the `plotDeletionsByVpooled` can be used. 
However, it is recomended to use this function for multiple individuals.
```{r, eval=TRUE, warning=FALSE,fig.height=4,fig.width=8}
# Plot the deletion heatmap
plotDeletionsByVpooled(del_db)
```


### Infering double chromosome deletion by relative gene usage

....

The `deletionsByBinom` function uses the V pooled approach to detect single chromosomal deletion for D and J.


```{r, eval=TRUE, warning=FALSE}
# Infered deletion summary table
del_binom_db <- deletionsByBinom(samples_db)
head(del_binom_db)
```

## References

1. [Gidoni *et al.* (2018)](https://doi.org/10.1101/314476).
1. [Alamyar *et al.* (2010)](http://www.imgt.org/IMGTindex/IMGTHighV-QUEST.html)
2. [Munshaw and Kepler (2010)](http://www.ncbi.nlm.nih.gov/pubmed/20147303)
3. [Lefranc *et al.* (2003)](http://www.ncbi.nlm.nih.gov/pubmed/12477501)

[1]: https://doi.org/10.1101/314476 "Gidoni et al. (2018)"
[2]: http://www.imgt.org/IMGTindex/IMGTHighV-QUEST.html "Alamyar et al. (2010)" 
[3]: http://www.ncbi.nlm.nih.gov/pubmed/20147303 "Munshaw and Kepler (2010)" 
[4]: http://www.ncbi.nlm.nih.gov/pubmed/12477501 "Lefranc et al. (2003)"

