---
title: "RAbHIT: R Antibody Haplotype Infrence Tool"
author: "Moriah Gidoni, Ayelet Peres, Gur Yaari"
date: "Last modified `r Sys.Date()`"
output:
  html_document:
    fig_height: 20
    fig_width: 15
    highlight: pygments
    theme: readable
    toc: yes
    includes:
      in_header: extLogo.html
  pdf_document:
    dev: pdf
    fig_height: 20
    fig_width: 15
    highlight: pygments
    toc: yes
    includes:
      in_header: header.tex
  md_document:
    fig_height: 20
    fig_width: 15
    preserve_yaml: no
    toc: yes
bibliography: bibliography.bib
csl: ieee-with-url.csl
link-citations: yes
urlcolor: blue
geometry: margin=1in
fontsize: 11pt
vignette: >
  %\VignetteIndexEntry{Haplotype infrence} 
  %\usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown} 
---
  
## Introduction 
  
Analysis of antibody repertoires by high throughput sequencing is of major importance in understanding adaptive immune responses. 
Our knowledge of variations in the genomic loci encoding antibody genes is incomplete, mostly due to technical difficulties in aligning short reads to these highly repetitive loci. 
The partial knowledge results in conflicting *V-D-J* gene assignments between different algorithms, and biased genotype and haplotype inference. Previous studies have shown that haplotypes can be inferred by taking advantage of *IGHJ6* heterozygosity, observed in approximately one third of the population [@kidd_inference_2012;@kirik2017parallel].

Here we provide a robust novel method for determining *V-D-J* haplotypes by adapting a Bayesian framework, **RAbHIT**. 
Our method extends haplotype inference to *IGHD* and *IGHV* based analysis, thereby enabling inference of complex genetic events like deletions and copy number variations in the entire population. It calculates a Bayes factor, a number that indicates the certainty level of the inference, for each haplotyped gene. 

More details can be found here:
  
[Gidoni, Moriah, et al. "Mosaic deletion patterns of the human antibody heavy chain gene locus as revealed by Bayesian haplotyping." bioRxiv (2018): 314476.](https://doi.org/10.1101/314476)

## Installation


For now, only the development version of `RAbHIT` is available for installation.
To install the package from the source code please use the code below:

```R
library(devtools)
install_bitbucket("yaarilab/rabhit")
```

For further installion instruction please refer to [https://bitbucket.org/yaarilab/rabhit](https://bitbucket.org/yaarilab/rabhit)

## Input

RAbHIT requires two main inputs:
  
1. Pre-processed antibody repertoire sequencing data with heterozygosity in at least one gene
2. Database of germline gene sequences

Antibody repertoire sequencing data is in a data frame format. Each row represents a unique 
observation and columns represent data about that observation. The names of the required columns are provided below along with a short description.

Column Name           | Description 
----------------------|---------------------------------------------------------
`SUBJECT`             | Subject name
`V_CALL`              | (Comma separated) name(s) of the nearest *V* allele(s) (IMGT format)
`D_CALL`              | (Comma separated) name(s) of the nearest *D* allele(s) 
`J_CALL`              | (Comma separated) name(s) of the nearest *J* allele(s) 


An example dataset is provided with the `rabhit` package. It contains unique naive b-cell
sequences, from a single individual.

The database of germline sequences should be provided in FASTA format with 
sequences gapped according to the IMGT numbering scheme[@lefranc2003imgt]. IGHV alleles in
the IMGT database (build 2014-08-4) are provided with this package. (object name)

```{r, eval=TRUE, message=FALSE, warning=FALSE}
library(rabhit)
# Load example sequence data and example germline database
data(sample_db, HVGERM, HDGERM)
```
### Pre-processing of the data

To get the most reliable result we suggest to follow the data pre-processing steps below.

1. Discover novel alleles (TIgGER[@gadala2018identification])
2. Infer genotype and reassign sequences accordingly (TIgGER[@gadala2018identification]) 
3. Sequence filtration by cells type:
    * For naive cells sequences, use only *V* genes with <=3 mutation and no mutation in *D* gene.
    * For PBMC cells sequences, first cluster sequences into clones (SHazaM[@gupta2015change]) then chose from each clone a representative sequence with the least number of mutations.
4. Preferably filter out non-functional sequences.


## Running RAbHIT

The functions provided by this package can be used to perform any combination of
the following:
  
1. Infer haplotype by anchor gene
2. Infer *D/J* single chromosome deletions by the *V* pooled approach
3. Infer double chromosome deletion by relative gene usage
4. Graphical output of the inferred haplotype 
5. Graphical output of the inferred deletion


### Infer haplotype by anchor gene

An individual's haplotype can be inferred using the functions `createFullHaplotype`.
The function infers the haplotype based on the provided anchor gene. Using this function a contingency table is created for each gene, *from which strand is inferred for each allele*.
The user can set the anchor gene for haplotyping as well as the column for which a haplotype should be inferred.

Prior to haplotyping, it is recommended to run TIgGER on the data, to detect new alleles and construct a genotype (TIgGER[@gadala2018identification]).

```{r, eval=TRUE, warning=FALSE}
# Inferred haplotype summary table
haplo_db <- createFullHaplotype(sample_db, toHap_col=c("V_CALL","D_CALL"),
                                hapBy_col="J_CALL", hapBy="IGHJ6", 
                                toHap_GERM=c(HVGERM, HDGERM))
```

```{r, eval=FALSE, warning=FALSE}
head(haplo_db)
```

```{r, eval=TRUE, warning=FALSE,echo=FALSE}
head(haplo_db,3)
```

```{r, eval=TRUE, warning=FALSE,fig.width=15,fig.height=10}
# Plot the haplotype map
plotHaplotype(haplo_db)
```

To further compare the haplotype inference between multiple individuals, the `hapHeatmap` function can be used. This function generate a heatmap of the alleles genes inferred for each chromosome and.

```{r, eval=TRUE, warning=FALSE, fig.width=12, fig.height=8}
# Load example sequence data
data(samples_db)
# Inferred haplotype summary table
haplo_db <- createFullHaplotype(samples_db, toHap_col=c("V_CALL","D_CALL"),
                        hapBy_col="J_CALL", hapBy="IGHJ6", toHap_GERM=c(HVGERM, HDGERM),
                        supress_print = T)
# Plot the haplotype inferred heatmap
hapHeatmap(haplo_db)
``` 

### Infering double chromosome deletion by relative gene usage

Gene usage tends to vary between individuals, in certain cases the relative gene usage of certain individuals are much lower than the rest of the population. To asses whether the low frequency arise from a deleted gene, a binomial test described in  Gidoni *et al.* (2018)[@gidoni2018mosaic] was implemented. They checked whether a certain relative gene usage of an individual is lower than the a chosen cutoff, for example for the *IGHV* genes, the chosen cutoff was $0.001$. The `deletionsByBinom` function implements the binomial and return the detect gene deletion for a certain individual.


```{r, eval=TRUE, warning=FALSE}
# Inferred deletion summary table
del_binom_db <- deletionsByBinom(samples_db)
head(del_binom_db)
```

For visualizing the deletion detected by `deletionsByBinom`, the `plotDeletionsByBinom` can be used. It is recommended to use this function for multiple individuals.

```{r, eval=TRUE, warning=FALSE,fig.height=9,fig.width=15}
# Don't plot IGHJ
del_binom_db <- del_binom_db[grep('IGHJ',del_binom_db$GENE, invert = T),]
# Inferred deletion summary table
plotDeletionsByBinom(del_binom_db) 
```


The detection of double chromosome deletion from the function can be then used for the haplotype inference, this prior knowledge of deletion can rise the certainty level of the inference of the genes where a deletion where a deletion was detected. The `createFullHaplotype` receives a vector of the deleted genes detected.
V gene labels marked in red represent low expressed genes for which deletions are inferred with low certainty. D gene labels marked in purple represent indistinguishable genes due to high sequence similarity, therefore alignment call is less reliable.


### Interactive Haplotype infrence with double chromosome deletions

As mentioned above the double chromosome deletions can be used within the haplotype inference. First the double deletion summary table should be generated with `deletionsByBinom`. Next, the haplotype can be inferred imputing the deletion summary table to the `deleted_genes` flag in the `createFullHaplotype` function.


```{r, eval=TRUE, warning=FALSE}
# Inferred deletion summary table
del_binom_db <- deletionsByBinom(sample_db)
# Input the summary table to the deleted_genes flag
haplo_db <- createFullHaplotype(sample_db, toHap_col=c("V_CALL","D_CALL"),
                        hapBy_col="J_CALL", hapBy="IGHJ6", toHap_GERM=c(HVGERM, HDGERM),
                        deleted_genes = del_binom_db, supress_print = T)
```

Another way to visualize the inferred haplotype is using the interactive HTML5 plot. To plot the HTML5 visualization, the `plotHaplotype` function should be used with the `html_output` flag marked `TRUE`. The function `saveWidget` from the `htmlwidgets` package can be used to save the interactive plot to a html file.    

```{r, eval=TRUE, warning=FALSE}
# Generate interactive haplotype plot
p <- plotHaplotype(haplo_db, html_output = TRUE)

# Save plot to html output
htmlwidgets::saveWidget(p, "haplotype.html", selfcontained = T)

# Plot the haplotype
p
```


### Haplotype inference deletion heatmap

For a group of individuals, the deletions detected by the haplotyping process can be visualized with `deletionHeatmap`.
The function create a heatmap of the deletion inferred and colors them by the certainty level ($lK$).

```{r, eval=TRUE, warning=FALSE,fig.height=12,fig.width=15}
# Inferred haplotype summary table for multiple subjects
haplo_db <- createFullHaplotype(samples_db, toHap_col=c("V_CALL","D_CALL"), 
                                hapBy_col="J_CALL", hapBy="IGHJ6", 
                                toHap_GERM=c(HVGERM, HDGERM), supress_print = T)
# plot deletion heatmap
deletionHeatmap(haplo_db)
```


### Infering D/J single chromosome deletion by V pooled approach

Since *V* gene heterozygosity is extremely common,
using *V* genes as anchors for haplotype inference could dramatically 
increase the number of people for which *D* haplotype can be inferred.
However, reliable haplotype inference using *V* genes as 
anchors requires a much greater sequencing depth than haplotype inference using J6 gene as an anchor.

The RAbHIT package offers a solution to  overcome the low number of sequences that connect a given *V-D* allele pair.
The package function applies an aggregation approach, in which information from several *V* heterozygous genes can be combined to infer *D* gene
deletions.

The `deletionsByVpooled` function uses the *V* pooled approach to detect single chromosomal deletion for *D* and *J*.


```{r, eval=TRUE, warning=FALSE}
# Inferred deletion summary table
del_db <- deletionsByVpooled(samples_db)
head(del_db)
```

For visualizing the deletion detected by `deletionsByVpooled`, the `plotDeletionsByVpooled` can be used. 
However, it is recommended to use this function for multiple individuals.
```{r, eval=TRUE, warning=FALSE,fig.height=4,fig.width=8}
# Plot the deletion heatmap
plotDeletionsByVpooled(del_db)
```

## Contact

For help, questions, or suggestions, please contact:

* [Moriah Gidoni](mailto:moriah.cohen@biu.ac.il)
* [Ayelet Peres](mailto:peresay@biu.ac.il)
* [Gur Yaari](mailto:gur.yaari@biu.ac.il)
* [Issue tracker](https://bitbucket.org/yaarilab/rabhit/issues?status=new&status=open)


## References

[1]: https://www.ncbi.nlm.nih.gov/pubmed/22205028 "Kidd *et al.* (2012)"
[2]: https://www.ncbi.nlm.nih.gov/pubmed/28388445 "Kirik *et al.* (2017)" 
[3]: http://www.ncbi.nlm.nih.gov/pubmed/12477501 "Lefranc et al. (2003)"
[4]: https://tigger.readthedocs.io/en/0.3.1/ "Gadala-Maria and Gidoni *et al.* (2018)" 
[5]: https://shazam.readthedocs.io/en/version-0.1.10/ "Gupta N *et al.* (2015)"
[6]: https://doi.org/10.1101/314476 "Gidoni et al. (2018)"
  
